### 2018-04-23 ident/docs/draft_3.tex
3rd draft based on prof.m's comments on draft 2

@@

	\section{Results:}
	First, in section \ref{sec:example}, we demonstrate the use of the methodology that we described in section \ref{sec:ident} to practically identify parameters in flux $v_1$ of the small gluconeogenic network (Figure \ref{fig:network}) model given in section \ref{sec:small-model} of the Supplementary Information. Identifiability results for fluxes $v_1$, $v_2$, $v_3$ and $v_5$ are also discussed in this section. We show that the ability to establish practical identifiability of kinetic parameters in metabolic network models relies upon the nonlinearity of the kinetic rate law formulations used to describe the fluxes. We also show that the degree of identifiability of the maximum reaction rate parameter is always higher than the degree of identifiability of the corresponding enzyme binding parameter irrespective of enzyme kinetic rate law used to describe the flux.

	Then, in section \ref{sec:experiment_type} we discuss the ability to determine the type of experiments for parameter identification for a given flux based on the informativeness of a given type of experiment. Here we show how the informativeness of a given type of experiment to identify a specific flux can be deduced from its contribution towards the practical identification of the parameters for a given flux.

	%In the following section we demonstrate our methodology to establish practical identifiability for one of the fluxes in this network (section \ref{sec:example}) (move to results).

	Then, in section \ref{sec:data_utility} we provide a motivation for the need for experimental design, especially to identify kinetic models of metabolic networks. We do this by looking at the number of combinations that can identify the maximum number of parameters. 
	In section \ref{sec:trends}

	Finally, we discuss some results arising out of the use of our methodology to determine the identifiability of parameters when data with additive noise is used in section \ref{sec:noise}.

	\subsection{Identifying parameters in kinetic models of metabolism: an example}\label{sec:example}	
	In this section, we illustrate the proposed methodology step by step to identify parameters of flux $v_1$ in the small metabolic network (Figure \ref{fig:network} and Section \ref{sec:small-model} of the Supplementary Information). We choose the expression for flux $v_1$ given in Equation (\ref{eq:flux1a}) for this demonstration. 
	\begin{align}\tag{S5}
	v_1 = V_1^{max}\frac{ac}{ac+K_{1}^{ac}(ne)}
	\end{align}		
	Since $\theta = \{V_1^{max}, K_1^{ac} (ne)\} \in \mathbb{R}^2$ for $v_1$, as mentioned in Section \ref{sec:experiments}, we need steady state concentration and flux measurements from at least two different experiments. So, from the $n_E = 21$ different experiments described in Section \ref{sec:experiments} and Table \ref{tab:pval}, we can choose multiple combinations of $p = 2$ experiments to satisfy the data requirements for identifying $v_1$ i.e., in Equation (\ref{eq:nonlineq}) $j = \{1, 2\}$. We label the available concentrations and fluxes as ${ac}^{(j)}$ and ${v_1}^{(j)}$, respectively. Then, the nonlinear algebraic equations shown in Equation (\ref{eq:nonlineq}) can be formulated for $v_1$ as:
	\begin{align*}%\label{eq:nonlin-flux2}
	{v_1}^{(j)} = V_{1}^{max}\frac{ac^{(j)}}{ac^{(j)}+K_{1}^{ac}(ne)} &&  j=\{1, 2\}
	\end{align*}

	Solving this simultaneous system of equations in $\mathbb{R}^2$ using Mathematica (Wolfram Research, USA), a computer algebra system, we get $p=2$ nonlinear algebraic expressions for parameters $V_1^{max}$ (Equation \ref{eq:v1_par}a) and $K_1^{ac}(ne)$ (Equation \ref{eq:v1_par}b). These expressions have the form shown in Equation (\ref{eq:theta-eq}).
	\begin{subequations}\label{eq:v1_par}
		\begin{align}		
		\theta_1 &= V_1^{max} = \frac{v_1^{(1)}v_1^{(2)}(ac^{(1)}-ac^{(2)})}{v_1^{(2)}ac^{(1)}-v_1^{(1)}ac^{(2)}}\\
		\theta_2 &= K_1^{ac}(ne) = \frac{ac^{(1)}ac^{(2)}(v_1^{(1)}-v_1^{(2)})}{v_1^{(2)}ac^{(1)}-v_1^{(1)}ac^{(2)}}
		\end{align}
	\end{subequations}
	In Equation (\ref{eq:v1_par}), the denominator of the right hand side expression is used to test the identifiability of parameters $V_1^{max}$ (Equation \ref{eq:v1_par}a) and $K_1^{ac}(ne)$ (Equation \ref{eq:v1_par}b) for different available experimental data combinations. Since the enzyme binding constant ($K_1^{ac}(ne)$) and the maximum reaction rate ($V_1^{max}$) cannot be negative, we can further constrain the criteria for identifiability for both these parameters by saying that the evaluated expressions in Equation (\ref{eq:v1_par}) should be non-negative (Figure \ref{fig:ident-flowchart}b). 
	We can also apply the proposed methodology to practically identify parameters in $v_1$ under the assumption that the protein concentration for the enzyme \textit{E} is available, in addition to the measured concentration of the metabolites and fluxes. In doing so, we get two expressions similar to the one shown in Equation (\ref{eq:v1_par}) for $k_1^{cat}$ and $K_1^{ac}$. Here, the value of $V_1^{max}$ in Equation (\ref{eq:flux1a}) above is substituted with $V_1^{max} = k_1^{cat}E$ instead. The globally identifiable parameter values for all fluxes modeled as Michaelis-Menten equations ($v_1$ and $v_2$) in provided in Supplementary Figure \ref{fig:parameter_value_proof}a and \ref{fig:parameter_value_proof}b.

	We apply the method described in Section \ref{sec:ident} and Figure \ref{fig:ident-flowchart}b, and illustrated above for $v_1$, to other fluxes of the network (Figure \ref{fig:network}) not described using Michaelis-Menten equations ($v_3$ and $v_5$). On applying this method to $v_3$, which is modeled using MWC kinetics, we find that we could not apply our methodology to determine closed form solution for all the parameters in $v_3$. However, when $v_3$ is instead modeled using Convenience kinetics, we find that the parameters are locally identifiable based on the definition we provide in Section \ref{sec:ident}. All three parameters $V_3^{max}$, $K_3^{fdp}$ and $K_3^{pep}$ used to model $v_3$ have two possible roots. We still could not determine identifiable values for these parameters (Supplementary Figure \ref{fig:parameter_value_v3}) using the constraints on the physical relevance of their values (see Section \ref{sec:ident} and Figure \ref{fig:ident-flowchart}). In the box plot in Supplementary Figure \ref{fig:parameter_value_v3}a and b we show all the possible estimated parameter values for both roots of all three parameters modeling $v_3$, when experimental data is generated using the MWC model for $v_3$. Due to the numerous estimated parameter values, we can designate these parameters are practically non-identifiable. Although using the Convenience kinetics model for $v_3$ to generate the in silico experimental data leads to a reduction in the uncertainty of the estimated parameters (Supplementary Figure \ref{fig:parameter_value_v3}c and d), we still end up with non-unique parameter estimates. This leads is to conclude that under the current conditions, flux $v_3$ is structurally locally identifiable and practically non-identifiable.

	However, $K_3^{fdp}$ and $V_3^{max}$ become globally identifiable when $K_3^{pep}$ is assumed to be known and fixed at a constant value, and $K_3^{pep}$ and $V_3^{max}$ become globally identifiable when $K_3^{fdp}$ is assumed to be known and fixed at a constant value (Supplementary Figure \ref{fig:parameter_value_v3_var}).

	Through Supplementary Figure \ref{fig:parameter_value_v3_var}, we also want to bring attention to the fact that parameter non-identifiability remains when there is a mismatch between the kinetic rate law used to generate the in silico data, and the kinetic rate law used to identify parameters. For example, in the specific case of $v_3$, we see that globally structurally identifiable parameters $V_3^{max}$ and $K_3^{fdp}$ (when $K_3^{pep}$ is fixed) are practically non-identifiable when data generated from experiments using the MWC kinetics is used to model $v_3$ (Supplementary Figure \ref{fig:parameter_value_v3_var}a). The same is true of globally structurally identifiable parameters $V_3^{max}$ and $K_3^{pep}$ when $K_3^{fdp}$ is fixed (Supplementary Figure \ref{fig:parameter_value_v3_var}b). Instead, when Convenience kinetics is used to generate in silico data, under the aforementioned conditions all the parameters are globally structurally and practically identifiable (Supplementary Figure \ref{fig:parameter_value_v3_var}c and d).

	Similarly, we applied the demonstrated methodology to identify flux $v_5$ that describes a transcription/translation reaction using Hill kinetics, based on only the available metabolomics and the fluxomics information. When we tried identifying the three parameters $V_e^{max}$, $K_e^{fdp}$ and $n_e$ in $v_5$, we could not obtain closed form solutions for these parameters using the computer algebra system. However, when we fixed the Hill coefficient $n_e$ as a known constant parameter, we were able to obtain closed form solutions for $V_e^{max}$ and $K_e^{fdp}$. Thus, similar to $v_3$ earlier, the dimension of the parameter space for $v_5$ had to be reduced from $\mathbb{R}^3$ to $\mathbb{R}^2$ to enable the application of the procedure that we have described in Section \ref{sec:ident}. In this instance, the identification required data from two different experiments. Again, just as we see for $v_3$, there are two different roots for $K_e^{fdp}$, making it locally structurally identifiable. On the other hand, $V_e^{max}$ has only one possible root, thereby it is a globally structurally identifiable parameter. However, only one of the two roots ($K_e^{fdp}(2)$) is practically identifiable based on the physical relevance criteria stated in Section \ref{sec:ident} and Figure \ref{fig:ident-flowchart}b. Thus, enforcing this criteria results in global practical identifiability of both $V_e{max}$ and $K_e^{fdp}$ in $v_5$ (Supplementary Figure \ref{fig:parameter_value_proof}c).

	In the following section, we describe and summarize the results on the identifiability of parameters in these fluxes ($v_1$, $v_2$, $v_3$ and $v_5$) based on the expressions and the identifiable roots we discussed in this section.

	\subsection{Degree of identifiability of different enzyme kinetic rate laws}\label{sec:ident_rate_laws}		
	In Figure \ref{fig:ident} and Supplementary Figure \ref{fig:ident_root2}, we show the number and percentage of experimental data combinations that are capable of identifying each parameter in each of the four fluxes $v_1$ (Figure \ref{fig:ident}a and e), $v_2$ (Figure \ref{fig:ident}b and f), $v_3$ (Figure \ref{fig:ident}c and g) and $v_5$ (Figure \ref{fig:ident}d and h) belonging to the network shown in Figure \ref{fig:network}. Based on the definition given in Section \ref{sec:degree_of_identifiability}, the percentages refer to the degree of identifiability of each parameter. We see from Figure \ref{fig:ident} and Supplementary Figure \ref{fig:ident_root2} that the degree of identifiability of the maximum reaction rates ($V_i^{max}$), whenever they are identifiable for each of the four fluxes, is greater than or equal to the degree of identifiability of the corresponding enzyme binding ($K_i$) constants, or the allosteric activation constant ($K_3^{pep}$), in the respective reaction rate law models. We make this observation despite the fact that the four fluxes are modeled using three different enzyme kinetic rate laws: $v_1$ and $v_2$ are modeled using the Michaelis-Menten rate law, $v_3$ is modeled using the Convenience kinetic rate law and $v_5$ is modeled as a Hill equation with inhibition. The fact that in silico experimental data is generated using different kinetic rate law models for $v_3$ (MWC in Figure \ref{fig:ident}a-d vs Convenience kinetics in Figure \ref{fig:ident}e-h) has no bearing on this observation. We accordingly surmise that the maximum reaction rate parameters are always more identifiable (as indicated by their higher degree of identifiability) than their enzyme binding constant counterparts, irrespective of the enzyme kinetic rate law used to model the corresponding flux.

	When data based on the MWC model for $v_3$ is used for identifying parameters, less that 50\% of all data combinations can identify $v_1$ (47.5\% and 40.0\% in Figure \ref{fig:ident}a). Even though the degree of identifiability of $K_1^{ac}(ne)$ is significantly different when data derived from the Convenience kinetics model for $v_3$ is used for identification (25\% in Figures \ref{fig:ident}e), the degree of identifiability is still less that 50\% for both parameters. In contrast, all available data combinations (from the MWC model and the Convenience kinetics model) satisfy the requirements for identifying $v_2$ (100\% in Figures \ref{fig:ident}b and f).

	The difference in the degrees of identifiability between $v_1$ (Figure \ref{fig:ident}a) and $v_2$ (Figure \ref{fig:ident}b) can be attributed to the ability of data from different combinations of experiments to satisfy the conditions determined for practical identifiability of that parameter. To elaborate, we showed in Equation (\ref{eq:v1_par}) and Section \ref{sec:example} that for a combination of any two experiments to be capable of identifying $V_1^{max}$ and $K_1^{ac}$ in $v_1$, the experiments must have distinct acetate concentrations as well as a different uptake flux $v_1$ between them. So, in this instance, recall that the ability of a combination of experiments to be useful is determined by the ability of the input change (changes in the input acetate concentration) to effect a change in the measured value of $v_1$. Similarly, the identification of parameters for $v_2$ (Figure \ref{fig:ident}b) requires the experiments to distinguish between values of both $v_2$ as well as \textit{pep}. We elaborate on the type and nature of experiments that are useful and informative for parameter identification in $v_1$ and $v_2$ later in Section \ref{sec:experiment_type}.

	The difference in the degree of identifiability of $v_1$ that comes about from the use of experimental data from different rate laws used to model $v_3$ (Figures \ref{fig:ident}a and e) can be again attributed to the usefulness of experimental data. However, the difference in data usefulness between similar experiments arising due to differences in the kinetic rate law used for $v_3$ could be due to the difference in the dynamics of the network represented by either of the two models. The interaction between different components of the network (Supplementary Figure \ref{fig:network}) as exemplified by the different models plays an important role in determining the dynamic and the steady state response of the network. The difference in dynamics in turn contributes to different experimentally observed values of metabolite concentrations and fluxes (Supplementary Figure \ref{fig:experimental_data_dist}), which in turn changes the usefulness of the given combination of datasets, and its subsequent effect of parameter identifiability.

	Nonetheless, we would like to note that these variations in the dynamics contributed by the difference between the rate laws can be resolved with more information, i.e., experimental proteomic data that provide protein concentrations for different experiments. We earlier saw that relaxing the assumption on the available experiment data, i.e., assuming the availability of protein concentration (\textit{E}), in addition to the metabolite concentrations (\textit{pep} and \textit{fdp}) and fluxes, results in removal of uncertainty and global identifiability of $v_1$ (Supplementary Figure \ref{fig:parameter_value_proof}). Recall that any predicted parameter value for both $k_1^{cat}$ and $K_1^{ac}$ matches with the exact known value irrespective of the model (MWC vs Convenience kinetics) from which experimental data is obtained when enzyme concentrations are included for identifiability (Supplementary Figure \ref{fig:parameter_value_proof}a).  Additionally, we also see an improvement in the degree of identifiability of the corresponding parameters $k_1^{cat}$ and $K_1^{ac}$ (Supplementary Figure \ref{fig:ident_root2}a and b) when enzyme concentrations are available.

	We would also like to point out that a difference in either the ability to estimate true parameter values, or the the identifiability of parameter modeling $v_2$ does not arise due to difference in the kinetic models used to obtain in silico experimental data. This could be because of the fact that any and all available information that is required for the identification of parameters in $v_2$ is encompassed in the steady state data used for identifiability analysis.

	However, unlike $v_2$, we see that the degree of identifiability of both locally identifiable roots of parameters in $v_3$ changes with the change in the data (MWC vs Convenience kinetics) used for the analysis (Figures \ref{fig:ident}c and g and Supplementary Figures \ref{fig:ident_root2}b and e). Despite the difference in the dynamics represented by the different data sets, and its impact on the usefulness of data sets for identifiability purposes that was explained earlier, we could also allude to the structural non-identifiability between the parameters in $v_3$ for the impact that it would have on the practical identifiability of the parameters using in silico data from different sources. In order to ascertain this, we plot the degrees of identifiability of the parameters $V_3^{max}$, $K_3^{fdp}$ and $K_3^{pep}$ under conditions when they are globally identifiable in Supplementary Figure \ref{fig:ident_v3_var}. To recall from Section \ref{sec:example}, $V_3^{max}$ and $K_3^{fdp}$ are globally identifiable when $K_3^{pep}$ is constant, and $V_3^{max}$ and $K_3^{pep}$ are globally identifiable when $K_3^{fdp}$ is known and fixed. Now that structurally non-identifiability between parameters has been alleviated, we see that the uncertainty introduced by the data obtained from the MWC model still results in a degree of identifiability of only about 5\% for $K_3^{fdp}$ (Supplementary Figure \ref{fig:ident_v3_var}a). However, when $K_3^{fdp}$ is held constant all MWC model-based experimental datasets contribute to local practical identifiability of parameters $V_3^{max}$ and $K_3^{pep}$ (Supplementary Figure \ref{fig:ident_v3_var}c), albeit with a bot of uncertainty (Supplementary Figures \ref{fig:ident_v3_var}b). This ties back with our earlier observation that these parameters are not globally practically identifiable from data sets obtained using the MWC model (Supplementary Figure \ref{fig:parameter_value_v3_var}a and b). However, we see from Supplementary Figures \ref{fig:ident_v3_var}c and d that the globally structurally identifiable parameters are also practically identifiable based on the data obtained from the Convenience kinetics model. This can be corroborated by the parameter values shown in Supplementary Figure \ref{fig:parameter_value_v3_var}c and d, where data from the Convenience kinetics model is used for analysis.

	We did not observe any differences in the degree of identifiability of parameters in $v_5$ between the analyses that used data from either the MWC model or the Convenience kinetics model for $v_3$ (Figure \ref{fig:ident}d and h). The identifiability of parameters based on the first root obtained from the computer algebra system are shown in Supplementary Figure \ref{fig:ident_root2}c and f. We use a similar reasoning to that we suggested for identifiability of $v_2$ above to justify our results.

	%In using the convenience kinetic rate law model for $v_3$ (Equation \ref{eq:flux3_convkin}) for identifying the parameters $V_3^{max}$, $K_3^{fdp}$ and $K_3^{pep}$, we found that each of the three parameters have two different closed-form expressions owing to the presence of a square root term in their solutions. Thus, as per the stated definition in section \ref{sec:ident}, the parameters for $v_3$, described using the convenience kinetics rate laws, are only locally identifiable due to the non-unique but finite number of possible solutions for the three parameters. The two possible solutions for $V_3^{max}$, $K_3^{fdp}$ and $K_3^{pep}$ are denoted by (1) and (2) following the respective parameter names throughout the rest of the document: $V_3^{max}(1)$, $K_3^{fdp}(1)$, and $K_3^{pep}(1)$ refer to the first set of solutions, and $V_3^{max}(2)$, $K_3^{fdp}(2)$, $K_3^{pep}(2)$ refer to the other set of solutions.

	\subsection{Rationale for a priori experimental design through practical parameter identification}\label{sec:data_utility}
	%In section \ref{sec:example}, we described how the practical identifiability of parameters in a kinetic model of a metabolic network depended on the enzyme kinetic rate law used to model the fluxes that constitute the network. In this section we look at the rationale for a priori experimental design to collect meaningful data for parameter estimation. 	
	%We also showed how the methodology that we propose can help in determining the nature and type of experiments that one needs to perform to identify and estimate parameters in kinetic models in section \ref{sec:experiment_type}. 	
	\textcolor{red}{(should this section moved to SI?)}Although the need for better parameter estimation methods is well researched, within the kinetic modeling community (\textcolor{red}{include references related to Banga papers on global optimization methods for parameter estimation, Rudiyanto Gunawan, Lennart Ljung and Heijnen papers here}), the need for experimental design methods to satisfy the data needs for parameter estimation has not gained enough traction (\textcolor{red}{include Rudiyanto Gunawan papers, D-optimal design using FIM here}). Hence, we find the need to stress the necessity for experimental design methodologies, in this case, tailored specifically for metabolic network models.

	We do this by showing how useful the multiple combinations of experimental data are towards identifying parameters for each flux $v_1$ (Figure \ref{fig:data_utility}a), $v_2$ (Figure \ref{fig:data_utility}b), $v_3$ (Figure \ref{fig:data_utility}c) and $v_5$ (Figure \ref{fig:data_utility}d). We express the usefulness of data from any given combination of experiments on the basis of the number of parameters that the data from the combination can identify. The higher the number of parameters a combination of experiments can identify, the greater is the usefulness of that combination of experiments. The percentage of all data combinations that can estimate a given number of parameters within a given flux is also shown in Figure \ref{fig:data_utility}.

	For $v_1$ (Figure \ref{fig:data_utility}a), data from about 50\% of the 240 combinations of experiments cannot identify any parameter used to model the flux ($V_1^{max}$ and $K_1^{ac}(ne)$). In contrast, data from only about 5\% of the experiments is useless towards identifying any parameter in $v_3$, while there are no useless combinations of experiments from which data cannot be used to identify either $v_2$, or $v_5$. These data are also in agreement with the degree of identifiability observed for parameters in all these fluxes (Figure \ref{fig:ident} and Supplementary Figure \ref{fig:ident_root2}). Thus, when it is possible to obtain steady state data from numerous experiments to estimate parameters for a small metabolic network, there is a very high possibility of performing experiments that could provide non-informative data that can result in parameter non-identifiability without a priori experimental design.

	We see that in instances where the identifiability of the fluxes differ based on the data set used (data from MWC model vs data from Convenience kinetics model), the utility of the data sets towards identifying parameters are changes. For instance, when data from the MWC model is used to identify parameters in $v_3$ (Figure \ref{fig:data_utility}c), less than 10\% of the data sets can identify all three parameters. Whereas, when data from the Convenience kinetics model is used for the same purpose, slightly more than 50\% of the combinations can identify all three parameters. Note that this observations should be considered in light of the structural local identifiability and the practical non-identifiability of these parameters that we mentioned earlier in Sections \ref{sec:example} and \ref{sec:ident_rate_laws}.

	Although one could argue based on earlier observations that uncertainty in the model from which data is obtained, and the structural non-identifiability of some of the parameters in the models of the aforementioned fluxes that there may not be a need for experimental design if in vivo model uncertainty and structural non-identifiability can be resolved, one needs to only look at Supplementary Figure \ref{fig:other_data_utility}a and d. In these figures we show the utility of multiple combinations of metabolite and protein concentration, and flux data towards identifying $k_1^{cat}$ and $K_1^{ac}$ in $v_1$ when data from the MWC model and Convenience kinetics model are used, respectively. We see that about 45\% of the 240 combinations of data from 21 different experiments cannot identify either one of the two parameters, when these parameters are globally structurally and practically identifiable.

	In addition to providing information on the fraction of experiments that are informative and those that are not informative, Figure \ref{fig:data_utility} and Supplementary Figure \ref{fig:other_data_utility} also informs the ease of identifying parameters for a single flux. In conjunction with the degrees of identifiability shown in Figure \ref{fig:ident} and Supplementary Figure \ref{fig:ident_root2}, we see that identifying $v_2$ is much easier than identifying $v_1$. Since minimization of resource utilization requires us to use the least number of experiments to identify both $v_1$ and $v_2$, this indicates the need to carefully choose combinations of experiments that can identify $v_1$. A careful consideration of experiments for identifying $v_2$ is however not a major concern since there is a very high probability that experiments chosen to identify $v_1$ will also identify $v_2$. A similar argument can also be for determining experiments that can identify $V_e^{max}$ and $K_e^{fdp}$ in $v_5$: since $V_e^{max}$ has an identifiability of 100\%, the choice of experiments to completely identify parameters in $v_5$ rests with experiments that can identify $K_e^{fdp}$.

	Accordingly, if we can establish a hierarchy of identifiable parameters and the corresponding fluxes they model, this section gives credence to the idea that a priori experimental design is necessary, even when all possible in vivo measurements can be obtained, to get the most informative data for parameter identification and estimation at minimum cost. Additionally, having the ability to see the contribution of each type of experiment towards identifiability would prove sufficient to decide on the total number and type of experiments that need to be performed to identify all fluxes in a kinetic model of metabolism. So, next we look at the contribution of different types of experiments towards identifying parameters in each flux in the small network.

	\subsection{Experimental requirements for identifying parameters depend on the position of the flux in the metabolic network}\label{sec:experiment_type}	
	In the previous sections, we showed how ability to structurally identify parameters in a kinetic model of metabolism is dependent on the kinetic rate law used to model a specific flux (Section \ref{sec:example}). We also discussed how the the informativeness of experiments used to practically identify parameters in kinetic models is not only dependent on the in silico rate law, but is also a function of the uncertainty in the experimental data brought about by the differences between the in silico rate law and the in vivo rate law from which data is obtained (Section \ref{sec:ident_rate_laws}). Subsequently, we also showed the need for a priori experimental design to generate informative data to identify parameters in kinetic models.

	In systems identification terminology, all the aforementioned observations regarding data requirements for parameters identification can be tied to selecting experiments that are persistently excitable for the flux being identified. In systems identification, any input signal should be rich or informative enough to guarantee full excitement of the dynamics of the system \parencite{Ljung1994}. Only information obtained from such changes in the input can be used to completely identify the system over its entire dynamic range. %Input signals that guarantee to fully excite the dynamics of the system are termed as persistently excitable signals. In linear systems, persistence of excitation of any input signal can be theoretically guaranteed. However, for nonlinear systems, like the metabolic network we deal with in this paper, persistence of excitation of input signals cannot be theoretically guaranteed, and should be assessed on a case by case basis. The lack of persistence of excitation requires the design of experiments to satisfy the data needs for complete system identification.

	In the case of identifying fluxes within a metabolic network individually, each flux is not independent from the other fluxes that form the metabolic network. Each flux is connected to one or more fluxes through the metabolites that not only act as substrates and products, but also as allosteric or transcriptional activators and inhibitors (e.g., effect of \textit{pep} on $v_3$, or the effect of \textit{fdp} on $v_5$ and consequently on $v_1$ in Figure \ref{fig:network}). Hence, careful consideration of experiments is a necessity so that the data acquired can satisfy conditions for practical identifiability for all parameters modeling a flux, and subsequently, all fluxes within a network \parencite{Heijnen2013}. So, in this section we discuss the dependence of the variation in the degree of identifiability for parameters of different fluxes from the point of view of the informativeness of experiments from which data is collected for identification.

	We demonstrated in section \ref{sec:example}, using $v_1$ and $v_2$ as examples, that the ability of data from a given combination of experiments to identify parameters in a given flux depends on its ability to satisfy the conditions determined for practical identifiability of that parameter. Although we were able to determine the requirements for identifying $v_1$ and $v_2$ based on the closed form expressions of the parameters, this may not be possible when a complex nonlinear rate law is used to model a flux. Using the Convenience Kinetic rate law to model $v_3$ is a case in point. So, in order to determine the type of experiments that help satisfy identifiability conditions for parameters of any flux modeled by any usable rate law, we look at the distribution of the different types of experiments in all data combinations that can identify the parameters within a flux. Recall that we use experimental data from five different types of experiments to test the practical identifiability of parameters in the model (Section \ref{sec:experiments} and Table \ref{tab:pval}).

	The contribution from different types of experiments towards identification of parameters in different fluxes is given in Figure \ref{fig:exp_info}. In Figure \ref{fig:exp_info}a and \ref{fig:exp_info}b, the contribution from different experiment types for identifying $V_1^{max}$ and $K_1^{ac}(ne)$ are given, respectively. The contribution of different experiments towards identifiability for $V_2^{max}$ (Figure \ref{fig:exp_info}c) and $K_2^{pep}$ (Figure \ref{fig:exp_info}d) are also shown. Note that the contributions are only shown for data sets obtained using the MWC model in Figure \ref{fig:exp_info}. Experiment type contributions towards identifiability based on data sets obtained from the Convenience kinetics model for $v_3$ are shown separately in Supplementary Figure .

	In agreement with what we discussed in Section \ref{sec:example}, i.e., experiments must be able distinguish between different values of $v_1$ and acetate, the contribution of experiments that involve changes in the acetate concentrations, which consequently bring about changes in the value of $v_1$, contribute to a significant part (> 50\%) of the identifiable experimental data combinations in comparison to the other types of experiments (Figure \ref{fig:exp_info}a and b). Since less than 50\% of all data combinations can satisfy these requirements, and can consequently identify $v_1$ (Figure \ref{fig:ident}a), we say that identifiability analysis is crucial to determine the minimum number of experiments, as well as the nature of experiments that can help identify parameters for $v_1$.

	In contrast to $v_1$, we see that both the enzyme perturbation and the acetate perturbation experiments have the same contribution towards combinations that can identify parameters for $v_2$ (Figure \ref{fig:exp_info}c and d). Hence, in conjunction with the high degree of identifiability seen in Figure \ref{fig:ident}b for $v_2$, we can deduce that almost all experiment types are persistently excitable for $v_2$ i.e., most experiments can bring about noticeable changes to both \textit{pep} and $v_2$. Consequently, this confirms our earlier hypothesis that in comparison to selecting experiments to identify $v_1$, there is very little restriction on the types of experiments that are informative enough to identify $v_2$.

>>>>

	\subsection{Kinetic model of gluconeogenesis in \textit{E. coli}}\label{sec:small-model}
	A previously proposed kinetic model \parencite{Kotte2014, Srinivasan2017} for acetate consumption through gluconeogenesis (Figure \ref{fig:network}) is used as a case study to illustrate identifiability analysis for experimental design for parameter estimation in kinetic models of metabolism. The kinetic model is described below.

	\begin{figure}[!tbhp]
		\centering{\includegraphics[width=.3\textwidth,height=.6\textheight,keepaspectratio]{figures/figure5/Figure1_NetworkA}}
		\caption{The previously published small metabolic network for gluconeogenesis used to demonstrate our practical identifiability method for kinetic models of metabolism.}\label{fig:network}
	\end{figure}

	\begin{equation}\label{eq:ode1}
	\frac{d}{dt}pep=v_1-v_2-v_4
	\end{equation}
	\begin{equation}\label{eq:ode2}
	\frac{d}{dt}fdp=v_2-v_3
	\end{equation}
	\begin{equation}\label{eq:ode3}
	\frac{d}{dt}E=v_5 - d E
	\end{equation}
	The kinetic expressions for fluxes $v_1$ through $v_5$ are given below. The consumption of acetate through $v_1$ and conversion of \textit{pep} through $v_2$ are expressed in Equations (\ref{eq:flux1}) and (\ref{eq:flux2}) respectively using Michaelis-Menten kinetics. The acetate flux through $v_1$ is also governed by the quantity of available enzyme E. 
	\begin{equation}\label{eq:flux1}
	v_1 = k_{1}^{cat}E\frac{ac}{ac+K_{1}^{ac}}
	\end{equation}		
	The model for flux $v_1$ of the small network (Figure \ref{fig:network}), uses the concentration of the enzyme E as a variable (Equation \ref{eq:flux1}). Since we assume that steady state experimental information is only available for metabolite concentrations and fluxes, and not for enzymes (again the details on relaxing this assumption are discussed later), the expression in Equation (\ref{eq:flux1}) for $v_1$ cannot be used for identifying parameters $k_1^{cat}$ and $K_1^{ac}$. So, we modify the Michaelis-Menten kinetic rate law expression to eliminate the enzyme concentration E as a variable in Equation (\ref{eq:flux1a}). Consequently $k_1^{cat}$ is replaced by $V_1^{max}$ as a parameter to describe $v_1$. The corresponding enzyme binding constant is denoted as $K_1^{ac} (ne)$ to distinguish it from the enzyme binding constant calculated in the presence of measured enzyme concentration data.
	\begin{align}\label{eq:flux1a}
	v_1 = V_1^{max}\frac{ac}{ac+K_{1}^{ac}(ne)}
	\end{align}		
	We choose the expression for flux $v_1$ given in Equation (\ref{eq:flux1a}) to demonstrate our method for practical identifiability. 	
	\begin{equation}\label{eq:flux2}
	v_2 = V_{2}^{max}\frac{pep}{pep+K_{2}^{pep}}
	\end{equation}
	\begin{equation}\label{eq:flux3}
	v_3 = V_{3}^{max}\frac{\tilde{fdp}\left(1+\tilde{fdp}\right)^3}{\left(1+\tilde{fdp}\right)^4+L_3\left(1+\frac{pep}{K_{3}^{pep}}\right)^{-4}}
	\end{equation}
	The allosterically regulated flux $v_3$ for the consumption of \textit{fdp} is expressed in Equation (\ref{eq:flux3}) using the Monod-Wyman-Changeux (MWC) model for allosterically regulated enzymes, where $\tilde{fdp}$ refers to the ratio of \textit{fdp} with respect to its allosteric binding constant $K_{3}^{fdp}$.

	The practically identifiability of parameters of a given flux are determined by solving a system of nonlinear algebraic equations using a computer algebra system (Section \ref{sec:ident}). We find that the nonlinearity of the MWC kinetic rate law used to model the allosteric regulation of $v_3$ makes it computationally intractable for determining the closed form expressions of the three parameters $V_3^{max}$, $K_3^{fdp}$ and $K_3^{pep}$ using a computer algebra system (Mathematica or SymPy in Python). In order to overcome this computational obstacle, we model the reaction rate for $v_3$ using the convenience kinetic rate law formulation \parencite{Liebermeister2006}. The corresponding expression obtained for $v_3$ is given below (Equation \ref{eq:flux3_convkin}). 	
	\begin{align}\label{eq:flux3_convkin}
	v_3 = V_3^{max}\left(\frac{1}{1 + \frac{K_3^{pep}}{pep}}\right)\left(\frac{\frac{fdp}{K_3^{fdp}}}{1 + \frac{fdp}{K_3^{fdp}}}\right)
	\end{align}

	The flux $v_4$ for the export of \textit{pep} is expressed as a linear equation dependent on $pep$ in Equation (\ref{eq:flux4}).
	\begin{equation}\label{eq:flux4}
	v_4 = V_{4}^{max}.pep
	\end{equation}		
	The production of enzyme E is represented by flux $v_5$. The inhibition of this flux by \textit{fdp} is modeled using Hill kinetics, where $K_e^{fdp}$ represents the Hill binding constant for the inhibiting metabolite \textit{fdp}, $n_e$ is the Hill exponent, and $V_e^{max}$ is the maximum reaction rate for $v_5$.
	\begin{align}\label{eq:flux5}
	v_5 = V_e^{max}\left(\frac{1}{1+\left(\frac{fdp}{K_{e}^{fdp}}\right)^{n_e}}\right)
	\end{align}

@@

	\section{Results:}
	First, in section \ref{sec:example}, we demonstrate the use of the methodology that we described in section \ref{sec:ident} to practically identify parameters in flux $v_1$ of the small gluconeogenic network (Figure \ref{fig:network}) model given in section \ref{sec:small-model}. In Section \ref{sec:proof} that follows, we show how the demonstrated methodology is capable of identifying and estimating parameters for fluxes $v_1$, $v_2$, and $v_5$ using steady state flux values and metabolite concentrations. We also discuss the ability of the proposed methodology to determine the structural and practical non-identifiability of parameters modeling $v_3$ in this section. The various ways in which this information can be used for designing experiments to generate data that can facilitate estimation of identifiable parameters are discussed in Section \ref{sec:experiment_type} that follows.

	\subsection{Identifying parameters in kinetic models of metabolism: an example}\label{sec:example}	
	\subsection{Relationship between structurally and practically parameter identifiability}\label{sec:proof}
	\subsection{Degree of identifiability of different enzyme kinetic rate laws}\label{sec:ident_rate_laws}			
	\subsection{Rationale for a priori experimental design through practical parameter identification}\label{sec:data_utility}	
	\subsection{Experimental requirements for identifying parameters depend on the position of the flux in the metabolic network}\label{sec:experiment_type}

	\subsection{Effect of noise on degree of identifiability is dependent on nonlinearity of enzyme kinetic rate law models}\label{sec:noise}
